---
title: "Effects of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Sean Tully - 10/23/2015"
output: pdf_document
fontsize: 6pt
geometry: margin=0.5in
---

# Overview #

The aim of this paper is to determine how vitamin C affects the overall tooth length in a randomized guinea pig population. Within the report the effects of delivery (supplement) and dosage will be reviewed.

## Data ##

The data for this analysis can be found in the r datasets library under ToothGrowth. Once loaded, the data can be parsed for easier calculation. [Code 1](#c01)

* 2 groups for supplement
* 3 groups for each combination of dosage pairs

``` {r, echo=FALSE}
library(datasets)
library(lattice)

data("ToothGrowth")
tg <- ToothGrowth
rm(ToothGrowth)
oj <- subset(tg, supp=="OJ")
vc <- subset(tg, supp=="VC")
d12 <- subset(tg, dose %in% c(0.5, 1.0))
d23 <- subset(tg, dose %in% c(1.0, 2.0))
d13 <- subset(tg, dose %in% c(0.5, 2.0))
```

A quick summary of the ToothGrowth data set reveals a collection of two distinct groups of sample data based on the type of supplement used. Both these groups indicate a general increase in guinea pig tooth length as the supplement dosage increase. [Code 2](#c02)

``` {r, echo=FALSE, fig.width=8, fig.height = 4}
xyplot(tg$len ~ tg$dose | tg$supp,xlab = "Dose",ylab = "Tooth Length", panel = function(x,y,...) {
    panel.xyplot(x,y,...)
    panel.lmline(x,y,col=1,lwd=2)
  })
    
```

## Tooth Growth Analysis ##

For the purposes of this analysis the interest is whether there is a significant difference between dosage amounts or supplement type. For each, confidence intervals will be calculated to numerically view if the null hypothesis can be rejected.

### Suppliment: ###

Looking only at which supplement was used to deliver vitamin C to the guinea pigs the confidence interval indicates that there is not a significant difference between methods. [Code 3](#c03)

``` {r, echo=FALSE}
library(knitr)
suppCI <- cbind(data.frame("Comparisons"=c("Suppliment")),rbind(
  t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = tg)$conf
)
)
names(suppCI) <- c("Comparisons","Lower Limit","Upper Limit")
kable(suppCI,digits=3)
```

### Dosage: ###

Ignoring the delivery system (Supplement) the effectiveness can be analyzed by comparing the effectiveness of each group by the others. [Code 4](#c04)

* Group1: Dose 0.5 vs 1.0
* Group1: Dose 1.0 vs 2.0
* Group1: Dose 0.5 vs 2.0

``` {r, echo=FALSE}
library(knitr)
doseCI <- cbind(data.frame("Comparisons"=c("Dose 1-2","Dose 2-3","Dose 1-3")),rbind(
  t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = d12)$conf,
  t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = d23)$conf,
  t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = d13)$conf
)
)
names(doseCI) <- c("Comparisons","Lower Limit","Upper Limit")
kable(doseCI,digits=3)
```

Here with in each coupling there is a significant difference between lower dosages and higher ones indicated by the confidence interval residing wholly in the negative region.

## Conclusions ##

* There is no distinctive difference between supplement provided the subjects
* Dosage level increase leads to a significant increase in tooth length

## Assumptions ##

* A random sampling of physically similar guinea pigs were used
* Measurements were taken from independent populations (unpaired)
* The variance between each test group is different

\pagebreak

# Appendix: 

##### c01 

Import and Parse Dataset:

``` {r, eval=FALSE}
          library(datasets)
          library(lattice)
          
          data("ToothGrowth")
          tg <- ToothGrowth
          rm(ToothGrowth)
          oj <- subset(tg, supp=="OJ")
          vc <- subset(tg, supp=="VC")
          d12 <- subset(tg, dose %in% c(0.5, 1.0))
          d23 <- subset(tg, dose %in% c(1.0, 2.0))
          d13 <- subset(tg, dose %in% c(0.5, 2.0))
```

##### c02 

Exploritory lattice:

``` {r, eval=FALSE}
          xyplot(tg$len ~ tg$dose | tg$supp,xlab = "Dose",ylab = "Tooth Length", panel = function(x,y,...) {
              panel.xyplot(x,y,...)
              panel.lmline(x,y,col=1,lwd=2)
            })
```

##### c03 

Tabulate confidence intervals (Supplement):

``` {r, eval=FALSE}
          library(knitr)
          suppCI <- cbind(data.frame("Comparisons"=c("Suppliment")),rbind(
            t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = tg)$conf
          )
          )
          names(suppCI) <- c("Comparisons","Lower Limit","Upper Limit")
          kable(suppCI,digits=3)
```

##### c04 

Tabulate confidence intervals (Dosage):

``` {r, eval=FALSE}
          library(knitr)
          doseCI <- cbind(data.frame("Comparisons"=c("Dose 1-2","Dose 2-3","Dose 1-3")),rbind(
            t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = d12)$conf,
            t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = d23)$conf,
            t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = d13)$conf
          )
          )
          names(doseCI) <- c("Comparisons","Lower Limit","Upper Limit")
          kable(doseCI,digits=3)
```

